\documentclass[11pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{amssymb,amsmath,amsthm,longtable}
\usepackage{epsfig}
\usepackage{epsf}
\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}

% more packages
%\usepackage{draftcopy}
%\usepackage{aas_macros}
\numberwithin{equation}{section}
\usepackage{hyperref}

% draft watermark
\usepackage{draftwatermark}
\SetWatermarkText{DRAFT}
\SetWatermarkColor[gray]{0.9}
\SetWatermarkScale{7}

% begin equation, itemize, etc.
\def\be{\begin{equation}}
\def\ee{\end{equation}}
\def\bi{\begin{itemize}}
\def\ei{\end{itemize}}
\def\ben{\begin{enumerate}}
\def\een{\end{enumerate}}
\def\i{\item{}}
\def\edth{\check\partial}

% special math fonts
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\mb}[1]{\mathbf{#1}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\mc}[1]{\mathcal{#1}}
\newcommand{\mf}[1]{\mathfrak{#1}}
\newcommand{\mr}[1]{\mathrm{#1}}
\newcommand{\ms}[1]{\mathsf{#1}}
\newcommand{\mat}[1]{\mathsf{#1}}
\newcommand{\unit}{\mathds{1}}
\newcommand{\zero}{\mb{0}}
 
% some more definitions
\def\cross{\times}
\def\del{\nabla}
\def\grad{\vec\nabla}
\def\div{\grad\cdot}
\def\curl{\grad\cross}
\def\I{I\!\!\!-}
 
\def\defn{\underline{Definition}:\ }
\def\ex{\underline{Example}:\ }
\def\exer{\underline{Exercise}:\ }
\def\soln{\underline{Solution}:\ }
\def\theor{\underline{Theorem}:\ }
\def\pf{\underline{Proof}:\ }
\def\ques{\underline{Question}:\ }
\def\ans{\underline{Answer}:\ }
 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

\title{Searches for stochastic gravitational-wave backgrounds}
\author{Joseph D.\ Romano}
\date{Les Houches Summer School\\
July 2018}
\maketitle

\abstract{These lecture notes provide a brief introduction 
to detection methods used to search for a stochastic 
background of gravitational radiation---a superposition
of gravitational-wave signals either too weak or too 
numerous to individually detect.
The lectures are divided into two main pieces:
(i) an overview, consisting of a description of different 
types of gravitational-wave backgrounds and an introduction 
to the correlation method using multiple detectors;
(ii) details, extending the previous discussion to 
non-trivial detector response, what to do in the absence
of correlations, and a recently proposed Bayesian method 
to search for the gravitational-wave background produced 
by stellar-mass binary black hole mergers throughout the 
universe.
Some suggested exercises for the reader are given 
throughout the text.
\pagebreak

\tableofcontents
\pagebreak

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% noindent, space between paragraphs
\setlength{\parindent}{0pt}
\setlength{\parskip}{\medskipamount}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Motivation}

A stochastic background of gravitational radiation 
is a superposition of gravitational-wave signals either 
too weak or too numerous to individually detect.
The individual signals making up the background are thus
{\em unresolvable}, unlike the large signal-to-noise 
binary black-hole (BBH) and binary neutron-star (BNS)
merger signals recently detected by the advanced LIGO 
and Virgo detectors.
But despite the fact that the individual signals are 
unresolvable, we shall see below that the detection of a 
stochastic gravitational-wave background (GWB) will 
be able to provide information about the statistical 
properties of the source.

\subsection{Comparison with the CMB}

The ultimate goal of gravitational-wave background
searches is to produce the GW analogue of Figure~\ref{f:CMB}.
Figure~\ref{f:CMB} is a sky map of the temperature 
fluctuations in the CMB blackbody radiation, relative 
to the $T_0 = 2.73~{\rm K}$ isotropic component.
(The dipole contribution due to our motion with respect 
to the cosmic rest frame has also been subtracted out.)
Recall that the CMB is a background of electromagnetic
radiation, produced at the time of last scattering,
roughly 380,000~yr after the Big Bang.
At that time, the universe had a temperature of 
$\sim 3000~{\rm K}$, roughly one thousand times larger 
than the temperature today, but cool enough for 
neutral hydrogen atoms to first form and photons to 
propagate freely.
The temperature fluctuations in the CMB radiation tell
us about the density of matter on the surface of last 
scattering and also about the integrated ...

For perspective, this map of the temperature fluctuations 
in the CMB was produced by the Planck mission in 2013,
almost 50 years after its initial detection by Penzias
and Wilson in 1965.
It took many years and many more improved experiments
(COBE, Boomerang, WMAP, Planck to name a few) to get to 
the high-precision measurements that we have now of the 
CMB.
So its somewhat sobering to realize that right now, 
in 2018, we have not yet detected even the isotropic 
component of the GWB.
So we have a long road ahead of us.  

\subsection{The background of BBH and BNS mergers in the 
LIGO band}

But fortunately, as mentioned above, the advanced LIGO
and Virgo detectors have detected other 
gravitational-wave signals from several individual BBH 
and BNS mergers.
These were very strong signals, having 
matched-filter signal-to-noise ratios $\gtrsim 10$, 
and false alarm probabilities $<2\times 10^{-7}$---i.e.,
5-sigma ``gold-plated" events.
Similar detections are expected in the upcoming 
observing run O3, which is scheduled to start in early 2019.

But we also expect many more signals, corresponding to
more distant mergers or smaller mass systems, which are 
individually undetectable (i.e., {\em subthreshold} events).
This weaker background of gravitational radiation is 
nonetheless detectable as a collectivity via the common 
influence of the gravitational waves on multiple detectors.

To get an idea of the statistical properties of this
GWB, we note that the local rate estimate from these 
first detections of BBH mergers in the LIGO band 
($9$-$240~{\rm Gpc}^{-3}\,{\rm yr}^{-1}$) lead to 
predictions of the total rate of such mergers throughout 
the universe to between $\sim\!1$~per minute to a few
per hour.
(\exer 1: Verify the above values for the total rate.)
Since the duration of BBH merger signals in band 
is $\sim\!1~{\rm s}$, the combined signal is 
highly-nonstationary (or {\em popcorn}-like).

One can do similar calculations for BNS mergers:
The predicted total rate for such events is roughly 
one event every 15~{\rm s}, while the duration of 
a BNS signal in band is roughly 100~{\rm s}. 
Thus, the signals overlap leading to a continuous
(or {\em confusion-limited}) background.

The combined signal from BBH and BNS mergers is potentially
detectable with advanced LIGO and Virgo, shortly after 
reaching design sensitivity (3-$\sigma$, correspond to a 
1/1000 false alarm probability).

This estimate of time to detection is based on 
the standard cross-correlation search, which assumes a
Gaussian-stationary background.

But there is a 
better method, proposed by Smith and Thrane, which we will
discuss at the end of these lectures.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Different types of stochastic backgrounds}

Stochastic backgrounds can be of either astrophysical 
or cosmological origin:

(i) A potential astrophysical background for the 
current generation of ground-based interferometers 
is the combined signal from the population of 
stellar-mass BBH and BNS mergers throughout the universe.
We will discuss the prospects of detecting this 
potential background throughout these lectures; the 
last section is devoted to a recently proposed 
detection method that targets this particular source.

(ii) A potential cosmological background is formed 
from {\em relic gravitational waves}---that is, 
quantum fluctuations in the geometry of space-time,
driven to macroscopic scales 
by a period of rapid expansion (e.g., inflation) 
a mere $\sim 10^{-32}~{\rm s}$ after the Big Bang.
This relic background is too weak to be detected by 
advanced LIGO, Virgo, etc.,
but is potentially detectable by its effect on the 
polarization of the cosmic microwave background (CMB)
radiation.
The Planck satellite and BICEP experiment (located
at the South Pole) are searching for this signal.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Mathematical characterization of a stochastic background}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Correlation methods - basic idea}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Some simple examples}

We now apply the above correlation method

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Non-trivial detector response}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Non-trivial correlations}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{What to do in the absence of correlations?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Searching for the background of binary black-hole
mergers}

\input refs

\end{document}

